# Review your R Training Prep

Prior to arriving in Brazzaville, we asked that you download R and RStudio and do some preparatory work. Let's work through some of that information together. 

#### Exercises {-}

**Use `R` like a calculator**  

**1.** Type a command in the *Console* to determine the sum of 596 and 198.  

**2.** Re-run the sum of 596 and 198 without re-typing it.  

**3.** Recall the command again, but this time adjust the code to find the sum of 596 and 298.  

**4.** Practice escaping an accidentally called command: recall your most recent command, then press the right key to clear the *Console*'s command line.  

&nbsp;  
**Recalling commands**  

**5.** Find the sum of the ages of everyone in your immediate family.  

**6.** Now recall that command and adjust it to determine the *average* age of the members of your family.  

**7.** Find the square root of *pi* and round the answer to the 2 decimal places.

&nbsp;    

**Finding errors**  

**8.** This line of code won't run; instead, `R` will wait for more with a `+` symbol. Find the problem and re-write the code so that it works.  

```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
5 * 6 +
```

**9.** The same goes for this line of code. Fix it, too.  

```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
sqrt(16
```

**10.** This line of code will trigger an error. Find the problem and re-write the code so that it works.

```{r,echo=TRUE,eval=FALSE,collapse=TRUE}
round(100/3,digits+3)
```

**11.** Type a command of your own into R that throws an error, then recall the command and revise so that `R` can understand it.

&nbsp;  


**Show that the following statements are TRUE:**

**12.** `pi` is greater than the square root of 9  

**13.** It is FALSE that the square root of 9 is greater than `pi`  

**14.** `pi` rounded to the nearest whole number equals the square root of 9  

&nbsp;  

**Asking `TRUE` / `FALSE` questions**

**15.** Write and run a line of code that asks whether these two calculations return the same result:

```{r,eval=FALSE,echo=TRUE,collapse=TRUE}
2*7 - 2*5 / 2
```

```{r,echo=TRUE,collapse=TRUE}
(2*7 - 2*5) / 2
```

**16.** Now write and run a line of code that asks whether the first calculation is larger than the second:
