
# Special Characters

#### Learning Goal {-}

* Learn how to print special characters using **`expression`** when adding text to plots in base `R`.

&nbsp; 

Often in science you want to use special characters in your plot axes' labels. Here you will learn how to print some of the more common special characters using `expression`.

&nbsp; 

**First set up an empty plot**

We do not need any data to practice this, so we'll use an empty plot, with no axes or axes labels.

Let's plot a single value, 1, and then tell R not to plot our single value using the command `type = "n"`. We will also suppress the axes using the command `xaxt="n"` and `yaxt="n"`, and we will leave the x and y labels blank, using `xlab=""` and `ylab=""`. 

```{r, echo=TRUE, eval=FALSE,collapse=TRUE}
plot(1,type = "n", xaxt="n", yaxt="n", xlab="", ylab="")
```

This gives us a blank plot with no data and no axes. If we wanted rid of the plot border too, we could just simply type `plot.new`

&nbsp; 

**Using `expression` to add special characters**

Let's imagine we have an experiment where we are measuring the response of soil CO~2~ emissions to temperature increases, measured in degrees Celsius (&deg;C).

Using the **`mtext`** command (short for "margin text"), we are going to label the y axis of our empty plot with "CO~2~" and we are going to label the x axis of our empty plot with the &deg; symbol.

Before we get to the special characters, let's quickly practice using `mtext` to see how it works.

Run this code:

```{r, echo=TRUE, eval=FALSE,collapse=TRUE}
mtext("Plot title", #This is the text you will add to your plot margin
        side=3,     #This tells R which margin you want to add the text to. Plot margins are numbered from the bottom, moving in a clockwise direction.
          line=1)   #Margins have different lines you can add text to. Increasing the line number moves the text further from the plot.
```

Can you see what `mtext` and its arguments are doing?


Now let's label our x and y axes with the required special characters. To do this we are going to have to use `expression` alongside `mtext`. 


Let's start with the y axis. To add CO~2~ to the y axis, we are going to have to tell R to make the 2 **subscript**. When using `expression` we can use **[ ]** to make text subscript. So our y axis label command will be:

```{r, echo=TRUE, eval=FALSE,collapse=TRUE}
mtext(expression(CO[2]),
      side=2,            #Side 2 for the y axis!
      line=1)
```


If we wanted instead to make something **superscript**, like if we wanted to report the R^2^ value on our plot, we'd use `^`, so it would read `expression(R^2)`.


Now we want to label our x axis with &deg;C. When using expression the &deg; symbol is simply **`degree`**. But we also want to add the letter C after the &deg; symbol. To do this we need to tell R to concatenate &deg; and C. To do this we need to use an \* . Normally in R \* represents multiplication, but when used with `expression` it represents **concatenate**. Therefore, to add our y label we use the following code:

```{r, echo=TRUE, eval=FALSE,collapse=TRUE}
mtext(expression(expression(degree*C)),
      side=1,            #Side 1 for the x axis!
      line=1)
```

&nbsp; 

**Concatenating special characters and normal text**

Now we want to create an empty plot where the y axis reads **"Temperature &deg;C"**.

Create an empty plot and this time try adding the word "Temperature" before the &deg; in your y axis `mtext` command.

What happens? 

You should get an error message. That's because if you want to add normal text to your `expression`, you need to use quotation marks (""). 

You also need to tell R to concatenate your normal text to your special character. To do that we need to use an \* . 

Therefore, this time try: 

`expression("Temperature " *degree*C)`

Does it work now? If so, good job!

&nbsp; 

Now create a plot where you add brackets around &deg;C, so that the y axis reads **"Temperature (&deg;C)"**.

This needs a bit of thought because the **( )** are normal text, wrapped around a special character, **&deg;C**. 

To achieve this, you need to incorporate the first bracket in the normal text containing "Temperature ": `"Temperature ("`. 

Then we concatenate our special character: `*degree*C`. 

Finally, you concatenate closing bracket, which is normal text:`*")"`

The full `expression` would therefore read:

`expression("Temperature ("*degree*C*")")`

&nbsp; 

`expression` also works directly in the `plot` function labels argument. So instead of using `mtext`, you could use:

```{r, echo=TRUE, eval=FALSE,collapse=TRUE}
plot(1, type = "n", xaxt="n", yaxt="n",
	   xlab=expression(CO[2]), 
		    ylab=expression("Temperature ("*degree*C*")"))
```

&nbsp; 

**Greek alphabet**

Other commonly required special characters in science/maths are characters from the Greek alphabet. These are very straight forward. You just type out the character name. 

Give it a try- create a new empty plot (`plot(1, type = "n", xaxt="n", yaxt="n", xlab="", ylab=""`) and using `mtext` and `expression` try adding the **`delta`** symbol to an axis. 

Now on a different axis, try capitalising the first letter, so that it reads **`Delta`** and see what happens. 

Try to think of some other Greek alphabet characters to try.

&nbsp; 

**Mathematical symbols**

Now let's look at some useful mathematical symbols. Create a new empty plot, and using `mtext` and `expression`, see what symbols are represented by the following code (maybe you can already guess some of them):

```
!=
%approx%
permil
infinity
```

Give these a try too (still using `expression`):

```
frac(x, y)
integral(f(x), a, b))
sqrt(x)
sum(x[i], i==1, n)
```
&nbsp; 

**Text Formatting**

And finally if you want to play around with formatting your text, give this a try...

`expression(bold("Hello, ")*italic("how's ")*underline("things?"))`




