---
title: R Training
site: bookdown::bookdown_site
output:
  bookdown::html_document2:
    includes:
      in_header: 
  bookdown::gitbook:
    includes:
      in_header: 
documentclass: book
github-repo: 
description: "A resource & workbook for learning R"
favicon: 
always_allow_html: true
---

```{r setup, include=FALSE, echo = FALSE}
# Basic knitr options
library(knitr)
opts_chunk$set(comment = NA, 
               echo=FALSE,
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE,
               fig.width = 9.64,
               fig.height = 5.9,
               fig.path = 'figures/')
options(scipen=999)
```

# Welcome to the course!

This won't be your standard classroom course where we formally teach you. Instead we will guide you through some things that we know and give you a chance to practice. Our aim is for you to feel confident from being able to write one line of code to calling yourself a proper researcher. 

If you have questions, ask. If you need someone to slow down or say it in a different way (comme le français), then say so. Because if you're thinking it, someone else probably is too. 

Follow along the next pages to get your computer up and running, and if you see a box, like the one below,

```{r, eval = FALSE, echo = TRUE}
be sure to type everything that you see in these boxes onto your own computer.
```

This course is about seeing **and doing**. Don't only read it. Do it.

Now that the formalities are out of the way... Buckle up, and let's learn how to work with data.